Setting parameters for biological models with ANIMO

### 2014-02-03 Paper.tex
Last revision: adjusted minor text problems, added a couple more references

\begin{abstract}
ANIMO (Analysis of Networks with Interactive MOdeling) is a software for modeling biological networks, such as e.g. signaling, metabolic or gene networks.
An ANIMO model is essentially the sum of a network topology (describing reactants and interaction types in form of a graph) and a number of interaction parameters. The topology describes the interactions between biological entities in form of a graph, while the parameters determine the speed of occurrence of such interactions.

@@

Proteins are normally expressed at different concentrations in different individuals of the same species, and yet the
overall behavior of their biological networks does not differ significantly. This phenomenon has led to the notion that biological
networks are inherently \emph{robust}~\cite{bio-robust1, bio-robust2}. In modeling terms, this means that,
if a model is a close representation of a biological network, most of the parameters of that model
can vary inside a certain interval without influencing the qualitative behavior of the whole network.
The interactive approach of ANIMO is based on the assumption of robustness, as our tool is mainly aimed at the
development of network models with a focus on the topology.
Ideally, the biologist can ``play'' with the topology of a network, working
more towards matching the qualitative behavior of experimental data, rather than precisely reproducing it.
However, it is not our intention to concentrate exclusively on the network topology:
in many cases a better parameter choice can improve the behavior of a network more than the addition of new components.
Indeed, making an unnecessarily complex model could reduce its usefulness both in terms of analysis
performances and closeness to reality. The first problem is simply due to the complexity of a network,
which would require more and more computational resources to be analysed\footnote{We also refer to the
problem of \emph{state space explosion}: when a model contains too many loosely coupled components,
the set of its possible evolutions grows exponentially, to the point of making it impossible to effectively apply analysis
techniques.}.
Realism of network models is more related to their
ultimate usefulness: a model that explains a particular behavior very well but contains many nodes marked
as ``unknown'' has little applicability, as its connection with known processes is very loose.
Therefore, it is desirable that a better parameter set is regularly sought for during the design cycle of
a complex biological network model. Some support for parameter choice was already provided in the first versions of ANIMO.
We present here an extended set of tools aimed at achieving a closer fit between ANIMO models and experimental data.
A guideline on how to use these tools to get the best results will also be presented as an ideal workflow.
Thanks to the better awareness on parameter choice gained through this new extension of ANIMO, the biologists
will be able to judge more easily which are the most promising topologies for a network, and thus drive
the experimental research more efficiently.

% - topology - parameters
% - problemi: small example network with parameter variations (also: some parts of a network are more sensitive to parameter variations than others)
\section{ANIMO models}
The starting point of ANIMO is the traditional static representation of biological networks,
which can be easily drawn and managed in softwares like Cytoscape~\cite{cytoscape}. Indeed, ANIMO was implemented
as a plug-in to Cytoscape, with the aim of adding dynamics to the static representation of biological
networks, and thus allow for analysis on the behavior of such networks. The user interface of ANIMO
can be seen in Figure~\ref{fig:animo-ui}: at the center is the Network panel, where the network
model is represented in the familiar nodes-edges form used in the domain of biology.
Models in ANIMO are \emph{activity-based}, in the sense that nodes have their activity as main
property, and interactions among nodes change the activity of their targets. The concept of activity
can be intended for example as a generic post-translational modification a molecule can undergo to change its
function. In the case of a kinase, the phosphorylated state is usually interpreted as active.
In the context of a gene network, the activity of a node standing for a gene represents its current
transcriptional status. A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity. ANIMO models are based on the formalism of \tas~\cite{timed-automata-alur-dill},
which is not directly shown to the user. Indeed, ANIMO was designed with the aim of enabling expert biologists
to formalize their knowledge without the need for additional mathematical training.

@@

Of the two main components of an ANIMO model, only the topology is immediately visible to the user in the
Network panel; the parameters are accessed by double clicking the arcs representing node-node interactions.
The dialog window that is shown for an interaction contains the details of the abstract reaction kinetic
describing the interaction, together with the current value of its parameter $k$ (see Fig.~\ref{fig:animo-edge-ui}).
The unique parameter associated to any interaction in ANIMO is used as a scale factor to make the modelled
reaction occur faster or slower. We also provide the user with pre-set values for
$k$, encouraging an initial qualitative assignment of reaction rates as ``slow'', ``fast'' and so forth.
This approach is based on the assumption that biological networks are inherently robust:
once an acceptable set of parameters is found, a more precise parameter search will generally have
little impact on the fitness of a model to a reference data set.
However, as can be seen in the small example in Figure~\ref{fig:small-example-feedback},
robustness does not imply that any parameter choice will do.
In that example, all parameters are initially set to \emph{medium}
(Fig.~\ref{fig:small-example-feedback}{\sf a}).
The peak in EGFR was obtained setting the value of $k$ for {\sf EGFR internalization}~$\dashv$~{\sf EGFR}
to \emph{fast} (Fig.~\ref{fig:small-example-feedback}{\sf b}). As this is not enough
to get a peak also for ERK, which remains constantly inactive, a lower value of $k$
for {\sf ERK phosphatase}~$\dashv$~{\sf ERK} can be tried. A first attempt with
\emph{slow} (Fig.~\ref{fig:small-example-feedback}{\sf c}) leads to a low peak,
which can be increased by further lowering that parameter to \emph{very slow}
(Fig.~\ref{fig:small-example-feedback}{\sf d}). Further adjustment of the parameters
and non-default scenario choices leads to the graph shown in Figure~\ref{fig:animo-ui}.
Network motifs such as the feedback loop shown in Figure~\ref{fig:small-example-feedback}
make models more dependent on parameter settings.
As biological networks tend to heavily rely on cross-talk, network topologies can rapidly become
complex. Some manual parameter fitting is currently necessary for the more complex ANIMO models.
This work is sometimes slow and error-prone, taking away some of the user-friendliness
for which ANIMO aims.
We will show how ANIMO has been improved to make parameter choice easier for the user.

@@

In ANIMO, the \emph{activity level} of a reactant is a discrete value representing
the ``active'' fraction of a molecular species. For example,
a phosphorylated kinase is usually considered to be active, while its non-phosphorylated state
is considered inactive. The activity level is expressed in ANIMO with an integer value between
0 (completely inactive) and a maximum value that represents complete activity. The maximum value can
be chosen by the user on a range between 1 and 100, determining the granularity (or precision)
with which the activity of the reactant is represented in the model. For example, an ANIMO node
with 2 levels of granularity (thus with maximum activity set to 1) can be used to represent a gene, which is considered as either
completely active or completely inactive. A more precise representation of the activity level
is necessary when we want to represent more complex dynamics, where also intermediate values are considered
important. The activity level of a node can be changed in an ANIMO model by way of activations
and inhibitions, which make it respectively rise and decrease. Each occurrence of an abstract
interaction changes the activity level of the downstream node by 1 step. For example,
consider the interaction MEK $\rightarrow$ ERK, which represents the activation of ERK by (active) MEK.
Each time that interaction occurs, the activity level of ERK rises by 1: if ERK has 100 activity levels,
it will take 100 interactions to take it from full inactivity to full activity. The rate
at which such interactions will occur is determined by the scenario and parameter of the
interaction.

The scenarios are used to choose which are the nodes to be taken into account
when computing the speed at which an interaction occurs.
\begin{description}
  \item[Scenario 1] the simplest of the three, this scenario approximates an interaction
    without taking into account the abundance of substrate: the activation/inhibition rate
    depends only on the activity level of the upstream node. In the MEK $\rightarrow$ ERK
    example, the rate of the interaction would be linearly dependent on the current activity level of MEK.
  \item[Scenario 2] scenario 2 models the interaction rate
    to be directly dependent both on the activity
    level of the upstream node and on the availability of substrate. By substrate we
    mean the inactive downstream node in case of activation interaction, and active downstream node in case of inhibition.
    In the above example, the interaction rate is linearly dependent on
    both the activity level of MEK and the inactive fraction of ERK.
    The inactive level of a node is computed by subtracting the current activity level from the maximum activity.
  \item[Scenario 3] the nodes on which this scenario depends can be chosen directly by the user,
    and can be either the active or inactive fraction of any two nodes in the network.
    This scenario can be used to represent an AND-gate, where two nodes are required to
    be simultaneously active in order to influence their target.
\end{description}

@@

By changing the title of a simulation, a user can keep track of different versions of their model,
which can be also kept and saved with the model, to be used in future sessions or for sharing purposes.
Should a user want to backtrack to a particular version of the network, the ``Reset to this'' button
can be used. As an additional help to recall what a model version consisted of, the tooltip of
the ``Reset to this'' button shows an image of the network topology, taken in the moment when
the selected simulation was begun. A click on the ``Reset to this'' button will reset the
network topology and parameters at the ones used to generate the selected simulation.
As the adaptation of a model may proceed on different paths, the ``Difference with\dots'' and
``Reset to this'' buttons provide the user with some help with tracking the changes to the model
and selecting the most promising ones.

@@

\begin{table}
\centering
\begin{tabular}{|l|l|}
\hline
 Reaction & $k$ \\
\hline
\hline
EGF $\rightarrow$ EGFR & Fast \\
EGFR $\rightarrow$ EGFR internalization & Medium \\
EGFR internalization $\dashv$ EGFR & Very Fast \\
EGFR $\rightarrow$ ERK & Fast \\
ERK phosphatase $\dashv$ ERK & Medium\\
\hline
\end{tabular}
\caption{Qualitative values of the parameters $k$ used in the example model to obtain the graph shown in Figure~\ref{fig:animo-ui}.\label{tab:parameters}}
\end{table}

@@

A simplified graphical representation of the workflow can be found in Figure~\ref{fig:workflow}: the three
degrees of precision with the parameter settings are highlighted with progressively darker shades of blue.
Moreover, the figure helps to stress the fact that in our method changes in topology (green boxes) should be made as a
last step, after a reasonable attempt at making the model fit via parameter adjustments.
Again, it is important to remember that if a model behaves correctly only for very narrow parameter settings
then it is not as robust as expected, so its topology may need to be reviewed.

@@

In the future, we plan to improve our parameter analysis techniques, possibly using some of
the techniques already developed for testing parameter sensitivity~\cite{inverse-ta1, inverse-ta2} and
robustness~\cite{robust-ta1, robust-ta2, robust-ta3, robust-ta4} in \tas\ and hybrid systems more in general~\cite{decidability-ta-ha, imitator2}.
As already stated, parameter sensitivity is very important in biological models: if a treatment
derived from a model is based on narrow bounds for a parameter, that parameter must not exhibit significant
variety among patients. By introducing an automatic analysis for parameter sensitivity, our tool
could point out the zones of a network model where the user should concentrate more, for example by
improving the topology, or by more focused parameter sweeps.
Moreover, we also aim at applying some results from automata learning to the generation of
\tas\ models based on experimental data: generating models that exhibit specific behaviors
implies a careful definition of their parameter space.
Finally, parametric model checking represents an interesting development direction:
having a tool automatically output the parameter intervals for which some
given outputs are observed is a highly valued resource for the biological researcher.
Investigating the properties of the specific \tas\ models used in ANIMO
could lead to important tractability results valid in our case, allowing us
to apply methods not valid for \tas\ in general.
%FUTURE WORK!!
%parameter sensitivity: se un trattamento dipende da un parametro specifico, e' il caso che non si registri troppa varieta' per quel parametro tra i pazienti.
% per quei parametri non sensibili, si eviti di concentrare troppi sforzi. Il tool potrebbe automaticamente determinare le ``zone calde'' della rete
%anche alla luce di una possibile generazione di modelli t.a. basati su dati sperimentali, trovare modi intelligenti di generare i parametri
%il fatto che i modelli che usiamo possano identificare un sottinsieme ristretto di t.a. potrebbe essere vantaggioso per la trattabilita' dei problemi
%parametrized model checking: con che parametri e' possibile raggiungere questo stato?
%definisci un obiettivo di quello che vuoi ottenere, e lui ti dice i [range di] parametri che arrivano a quell'evoluzione
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri

>>>>

\begin{abstract}
ANIMO (Analysis of Networks with Interactive MOdeling) is a software for modeling biological networks, such as e.g. signaling, metabolic or gene networks.
An ANIMO model is essentially the sum of a network topology and a number of interaction parameters. The topology describes the interactions between biological entities in form of a graph, while the parameters determine the speed of occurrence of such interactions.

@@

Proteins are normally expressed at different concentrations in different individuals of the same species, and yet the
overall behavior of their biological networks does not differ significantly. This phenomenon has led to the notion that biological
networks are inherently \emph{robust}~\cite{bio-robust1, bio-robust2}. In modeling terms, this means that,
if a model is a close representation of a biological network, most of the parameters of that model
can vary inside a certain interval without influencing the qualitative behavior of the whole network.
The interactive approach of ANIMO is based on the assumption of robustness, as our tool is mainly aimed at the
development of network models with a focus on the topology.
Ideally, the biologist can ``play'' with the topology of a network, working
towards matching the qualitative behavior of experimental data, rather than precisely reproducing it.
However, it is not our intention to concentrate exclusively on the network topology:
in many cases a better parameter choice can improve the behavior of a network more than the addition of new components.
Indeed, making an unnecessarily complex model could reduce its usefulness both in terms of analysis
performances and closeness to reality. The first problem is simply due to the complexity of a network,
which would require more and more computational resources to be analysed\footnote{We also refer to the
problem of \emph{state space explosion}: when a model contains too many loosely coupled components,
the set of its possible evolutions grows exponentially, to the point of making it impossible to effectively apply analysis
techniques.}.
Realism of network models is more related to their
ultimate usefulness: a model that explains a particular behavior very well but contains many nodes marked
as ``unknown'' has little applicability, as its connection with known processes is very loose.
Therefore, it is desirable that a better parameter set is regularly sought for during the design cycle of
a complex biological network model. Some support for parameter choice was already provided in the first versions of ANIMO.
We present here an extended set of tools aimed at achieving a closer fit between ANIMO models and experimental data.
A guideline on how to use these tools to get the best results will also be presented as an ideal workflow.
Thanks to the better awareness on parameter choice gained through this new extension of ANIMO, the biologists
will be able to judge more easily which are the most promising topologies for a network, and thus drive
the experimental research more efficiently.

% - topology - parameters
% - problemi: small example network with parameter variations (also: some parts of a network are more sensitive to parameter variations than others)
\section{ANIMO models}
The starting point of ANIMO is the traditional static representation of biological networks,
which can be easily drawn and managed in softwares like Cytoscape~\cite{cytoscape}. Indeed, ANIMO was implemented
as a plug-in to Cytoscape, with the aim of adding dynamics to the static representation of biological
networks, and thus allow for analysis on the behavior of such networks. The user interface of ANIMO
can be seen in Figure~\ref{fig:animo-ui}: at the center is the Network panel, where the network
model is represented in the familiar nodes-edges form used in the domain of biology.
Models in ANIMO are \emph{activity-based}, in the sense that nodes have their activity as main
property, and interactions among nodes change the activity of their targets. The concept of activity
can be intended for example as a generic post-translational modification a molecule can undergo to change its
function. In the case of a kinase, the phosphorylated state is usually interpreted as active.
In the context of a gene network, the activity of a node standing for a gene represents its current
transcriptional status.

A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity. ANIMO models are based on the formalism of \tas~\cite{timed-automata-alur-dill},
which is not directly shown to the user. Indeed, ANIMO was designed with the aim of enabling expert biologists
to formalize their knowledge without the need for additional mathematical training.

@@

Of the two main components of an ANIMO model, only the topology is immediately visible to the user in the
Network panel; the parameters are accessed by double clicking the arcs representing node-node interactions.
The dialog window that is shown for an interaction contains the details of the abstract reaction kinetic
describing the interaction, together with the current value of its parameter $k$ (see Fig.~\ref{fig:animo-edge-ui}).
The unique parameter associated to any interaction in ANIMO is used as a scale factor to make the modelled
reaction occur faster or slower. We also provide the user with pre-set values for
$k$, encouraging an initial qualitative assignment of reaction rates as ``slow'', ``fast'' and so forth.
This approach is based on the assumption that biological networks are inherently robust:
once an acceptable set of parameters is found, a more precise parameter search will generally have
little impact on the fitness of a model to a reference data set.
However, as can be seen in the small example in Figure~\ref{fig:small-example-feedback},
robustness does not imply that any parameter choice will do.
In that example, all parameters are initially set to \emph{medium}
(Fig.~\ref{fig:small-example-feedback}{\sf a}).
The peak in EGFR was obtained setting the value of $k$ for {\sf EGFR internalization}~$\dashv$~{\sf EGFR}
to \emph{fast} (Fig.~\ref{fig:small-example-feedback}{\sf b}). As this is not enough
to get a peak also for ERK, which remains constantly inactive, a lower value of $k$
for {\sf ERK phosphatase}~$\dashv$~{\sf ERK} can be tried. A first attempt with
\emph{slow} (Fig.~\ref{fig:small-example-feedback}{\sf c}) leads to a low peak,
which can be increased by further lowering that parameter to \emph{very slow}
(Fig.~\ref{fig:small-example-feedback}{\sf d}). Further adjustment of the parameters
and non-default scenario choices (explained later) leads to the graph shown in Figure~\ref{fig:animo-ui}.
Network motifs such as the feedback loop shown in Figure~\ref{fig:small-example-feedback}
make models more dependent on parameter settings.
As biological networks tend to heavily rely on cross-talk, network topologies can rapidly become
complex. Some manual parameter fitting is currently necessary for the more complex ANIMO models.
This work is sometimes slow and error-prone, taking away some of the user-friendliness
for which ANIMO aims.
We will show how ANIMO has been improved to make parameter choice easier for the user.

@@

In ANIMO models, the \emph{activity level} defines the fraction of a molecular
species that is considered active.
The activity level is expressed in ANIMO with an integer value between
0 (completely inactive) and a maximum value that represents complete activity. The maximum value can
be chosen by the user on a range between 1 and 100, determining the granularity (or precision)
with which the activity of the reactant is represented in the model. For example, an ANIMO node
with 2 levels of granularity (thus with maximum activity set to 1) can be used to represent a gene, which is considered as either
completely active or completely inactive. A more precise representation of the activity level
is necessary when we want to represent more complex dynamics, where also intermediate values are considered
important. The activity level of a node can be changed in an ANIMO model by way of activations
and inhibitions, which make it respectively rise and decrease. Each occurrence of an abstract
interaction changes the activity level of the downstream node by 1 step. For example,
consider the interaction MEK $\rightarrow$ ERK, which represents the activation of ERK by (active) MEK.
Each time that interaction occurs, the activity level of ERK rises by 1: if ERK has 100 activity levels,
it will take 100 interactions to take it from full inactivity to full activity. The rate
at which such interactions will occur is determined by the scenario and parameter of the
interaction.

The scenarios are used to choose which are the nodes to be taken into account
when computing the speed at which an interaction occurs.
\begin{description}
  \item[Scenario 1] the simplest of the three, this scenario approximates an interaction
    without taking into account the abundance of substrate: the activation/inhibition rate
    depends only on the activity level of the upstream node. In the MEK $\rightarrow$ ERK
    example, the rate of the interaction would be linearly dependent on the current activity level of MEK.
  \item[Scenario 2] scenario 2 models the interaction rate
    to be directly dependent both on the activity
    level of the upstream node and on the availability of substrate. By substrate we
    mean the inactive downstream node in case of activation interaction, and active downstream node in case of inhibition.
    In the MEK $\rightarrow$ ERK example, the interaction rate is linearly dependent on
    both the activity level of MEK and the inactive fraction of ERK.
    The inactive level of a node is computed by subtracting the current activity level from the maximum activity.
  \item[Scenario 3] the nodes on which this scenario depends can be chosen directly by the user,
    and can be the active or inactive fraction of any two nodes in the network.
    This scenario can be used to represent an AND-gate, where two nodes are required to
    be simultaneously active in order to influence their target.
\end{description}

@@

By changing the title of a simulation, a user can keep track of different versions of their model,
which can be also kept and saved with the model, to be used in future sessions or for sharing purposes.
Should a user want to backtrack to a particular version of the network, the ``Reset to here'' button
can be used (see Fig.~\ref{animo-ui}, buttons above the graph). As an additional help to recall what a model version consisted of, the tooltip of
the ``Reset to here'' button shows an image of the network topology, taken in the moment when
the selected simulation was begun. A click on the ``Reset to here'' button will reset the
network topology and parameters at the ones used to generate the selected simulation.
As the adaptation of a model may proceed on different paths, the ``Difference with\dots'' and
``Reset to here'' buttons provide the user with some help with tracking the changes to the model
and selecting the most promising ones.

@@

\begin{table}
\centering
\small
\begin{tabular}{|l|l|}
\hline
 Reaction & $k$ \\
\hline
\hline
EGF $\rightarrow$ EGFR & Fast \\
EGFR $\rightarrow$ EGFR internalization & Medium \\
EGFR internalization $\dashv$ EGFR & Very Fast \\
EGFR $\rightarrow$ ERK & Fast \\
ERK phosphatase $\dashv$ ERK & Medium\\
\hline
\end{tabular}
\caption{Qualitative values of the parameters $k$ used in the example model to obtain the graph shown in Figure~\ref{fig:animo-ui}.\label{tab:parameters}}
\end{table}

@@

A simplified graphical representation of the workflow can be found in Figure~\ref{fig:workflow}: the three
degrees of precision with the parameter settings are highlighted with progressively darker shades of blue.
Moreover, the figure helps to stress the fact that in our method changes in topology (green boxes) should be made as a
last step, after a reasonable attempt at making the model fit via parameter adjustments.
Again, it is important to remember that if a model behaves correctly only for very narrow parameter settings
then it may not be as robust as expected, so a review of its topology may be needed.

@@

In the future, we plan to improve our parameter analysis techniques, possibly using some of
the techniques already developed for testing parameter sensitivity~\cite{inverse-ta1, inverse-ta2} and
robustness~\cite{robust-ta1, robust-ta2, robust-ta3, robust-ta4} in \tas\ and hybrid systems more in general~\cite{decidability-ta-ha, imitator2}.
As already stated, parameter sensitivity is very important in biological models: if a treatment
derived from a model is based on narrow bounds for a parameter, that parameter must not exhibit significant
variety among patients. By introducing an automatic analysis for parameter sensitivity, our tool
could point out the zones of a network model where the user should concentrate more, for example by
improving the topology, or by more focused parameter sweeps.
Moreover, we also aim at applying some results from automata learning~\cite{test-based-modelling} to the generation of
\tas\ models based on experimental data: generating models that exhibit specific behaviors
implies a careful definition of their parameter space.
Finally, parametric model checking~\cite{param-mc} represents an interesting development direction:
having a tool automatically output the parameter intervals for which some
given outputs are observed is a highly valued resource for the biological researcher.
% Investigating the properties of the specific \tas\ models used in ANIMO
% could lead to important tractability results valid in our case, allowing us
% to apply methods not valid for \tas\ in general.
%FUTURE WORK!!
%parameter sensitivity: se un trattamento dipende da un parametro specifico, e' il caso che non si registri troppa varieta' per quel parametro tra i pazienti.
% per quei parametri non sensibili, si eviti di concentrare troppi sforzi. Il tool potrebbe automaticamente determinare le ``zone calde'' della rete
%anche alla luce di una possibile generazione di modelli t.a. basati su dati sperimentali, trovare modi intelligenti di generare i parametri
%il fatto che i modelli che usiamo possano identificare un sottinsieme ristretto di t.a. potrebbe essere vantaggioso per la trattabilita' dei problemi
%parametrized model checking: con che parametri e' possibile raggiungere questo stato?
%definisci un obiettivo di quello che vuoi ottenere, e lui ti dice i [range di] parametri che arrivano a quell'evoluzione
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri
